Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-02-10 paper/current/main.tex
add tables to main and supp text

\newcommand{\secref}[1]{Section~\ref{sec:#1}}
\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\tabref}[1]{Table~\ref{tab:#1}}
%\newcommand{\eqnref}[1]{\eqref{eq:#1}}
%\newcommand{\thmref}[1]{Theorem~\ref{#1}}
%\newcommand{\prgref}[1]{Program~\ref{#1}}
%\newcommand{\algref}[1]{Algorithm~\ref{#1}}
%\newcommand{\clmref}[1]{Claim~\ref{#1}}
%\newcommand{\lemref}[1]{Lemma~\ref{#1}}
%\newcommand{\ptyref}[1]{Property~\ref{#1}}

@@

We show that the embedding distance better predicts actual mobility flows than the geographic distance across three disparate datasets.
In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$).  Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d).
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores  calculated from the underlying mobility network (Fig.~\ref{fig:supp:distancemetrics}, Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
The embedding distance derived from neural embedding also explains more of the flux and better predicts mobility flows than simpler embedding baseline, such as distances derived from a singular-value decomposition on flow matrix and from a Laplacian Eigenmap embedding  \autocite{belkin2003laplacian} of the underlying data (Fig.~\ref{fig:supp:distancemetrics} and Fig.~\ref{fig:supp:distancermse}).
In sum, our results demonstrate that, consistently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

@@

@@

Neural embedding approaches offer a novel, data-driven solution for learning an effective and robust representation of locations based on trajectory data, encoding the complex and multi-faceted nature of mobility.
We demonstrated that a functional distance derived from the embedding can be used with the gravity model of mobility to better predict real-world mobility than does geographic distance.
Embedding distance outperformed geographic distance across distinct and disparate domains, including U.S. flight itineraries, Korean hotel accommodation reservations, and global scientific mobility.
We discover that this performance results from neural embeddings implicitly learning gravity law relationships between locations, making them particularly well suited to representing mobility.
Focusing on scientific mobility, we find that neural embedding's performance is driven by its ability to encode many aspects of scientific mobility into a single representation, including global and regional geography, shared languages, and prestige hierarchies, even without explicit information on these factors.

While we focus on three domains of mobility, this approach may be broadly applicable to other kinds of mobility data, such as general human migration, transit-network mobility, and more.
Moreover, this approach can be used to learn a functional distance even between entities for which no physical analog exists, such as between occupational categories based on individuals' career trajectories.
In addition to providing a functional distance that supports modeling and predicting mobility patterns, the structure of the neural embedding space is amenable to a range of unique applications for studying mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Embedding also allows us to quantitatively explore abstract notions such as academic prestige, and can potentially be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

@@

We calculate two forms of distance between locations.
The geographic distance, $g_{ij}$, is the pairwise geographic distance between locations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of locations.
In the case of U.S. flight itinerary and scientific mobility, we impute distance to 1 km when their distance is less than one kilometer.
In the case of Korean accommodation reservation data, because this data is intra-city mobility trajectory, we impute distance to 0.01 km when their distance is less than 0.01 km.
The embedding distance with the cosine distance, $d_{ij}$, is calculated as $d_{ij} = 1 - \frac{\bm{v}_{i} \cdot \bm{v}_{j}}{\lVert \bm{v}_{i} \rVert \lVert \bm{v}_{j} \rVert}$,
where $v_i$ and $v_j$ are the embedding vectors for locations $i$ and $j$, respectively.
Note that $d_{ij}$ is not a formal metric because it does not satisfy the triangle inequality.
Nevertheless, cosine distance is often shown to be useful in practice~\autocite{lerman2007embedding, brown1970migration, kim2018functional}.

>>>>

\newcommand{\secref}[1]{Section~\ref{sec:#1}}
\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\tabref}[1]{Table~\ref{tab:#1}}
\newcommand{\todo}[1]{{\leavevmode\color{orange}[TODO: #1]}}
%\newcommand{\eqnref}[1]{\eqref{eq:#1}}
%\newcommand{\thmref}[1]{Theorem~\ref{#1}}
%\newcommand{\prgref}[1]{Program~\ref{#1}}
%\newcommand{\algref}[1]{Algorithm~\ref{#1}}
%\newcommand{\clmref}[1]{Claim~\ref{#1}}
%\newcommand{\lemref}[1]{Lemma~\ref{#1}}
%\newcommand{\ptyref}[1]{Property~\ref{#1}}

@@

We show that the embedding distance better predicts actual mobility flows than the geographic distance across three disparate datasets.
In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$).  Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d).
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores  calculated from the underlying mobility network (Fig.~\ref{fig:supp:distancemetrics}, Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
The embedding distance derived from neural embedding also explains more of the flux and better predicts mobility flows than simpler embedding baseline, such as distances derived from a singular-value decomposition on flow matrix and from a Laplacian Eigenmap embedding  \autocite{belkin2003laplacian} of the underlying data (Fig.~\ref{fig:supp:distancemetrics} and Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}), in many cases even putperforming a direct factorization approach~\cite{levy2014neural}), which \textit{word2vec} efficiently approximates.
In sum, our results demonstrate that, consistently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

@@

\todo{add stuff about cosine distance performing better, SGD performing better}

@@

Neural embedding approaches offer a novel, data-driven solution for efficiently learning an effective and robust representation of locations based on trajectory data, encoding the complex and multi-faceted nature of mobility.
We demonstrated that a functional distance derived from the embedding can be used with the gravity model of mobility to better predict real-world mobility than does geographic distance.
Embedding distance outperformed geographic distance across distinct and disparate domains, including U.S. flight itineraries, Korean hotel accommodation reservations, and global scientific mobility.
We discover that this performance results from neural embeddings implicitly learning gravity law relationships between locations, making them particularly well suited to representing mobility.
Focusing on scientific mobility, we find that the embedding successfully encodes many aspects of scientific mobility into a single representation, including global and regional geography, shared languages, and prestige hierarchies, even without explicit information on these factors.

While we focus on three domains of mobility, this approach may be broadly applicable to other kinds of mobility data that is expected to follow a gravity-like pattern, such as general human migration, transit-network mobility, and more.
Moreover, this approach can be used to learn a functional distance even between entities for which no physical analog exists, such as between occupational categories based on individuals' career trajectories.
In addition to providing a functional distance that supports modeling and predicting mobility patterns, the structure of the neural embedding space is amenable to a range of unique applications for studying mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Embedding also allows us to quantitatively explore abstract notions such as academic prestige, and can potentially be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

@@

We calculate two main forms of distance between locations.
The geographic distance, $g_{ij}$, is the pairwise geographic distance between locations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of locations.
In the case of U.S. flight itinerary and scientific mobility, we impute distance to 1 km when their distance is less than one kilometer.
In the case of Korean accommodation reservation data, because this data is intra-city mobility trajectory, we impute distance to 0.01 km when their distance is less than 0.01 km.
The embedding distance with the cosine distance, $d_{ij}$, is calculated as $d_{ij} = 1 - \frac{\bm{v}_{i} \cdot \bm{v}_{j}}{\lVert \bm{v}_{i} \rVert \lVert \bm{v}_{j} \rVert}$,
where $v_i$ and $v_j$ are the embedding vectors for locations $i$ and $j$, respectively.
Note that $d_{ij}$ is not a formal metric because it does not satisfy the triangle inequality.
Nevertheless, cosine distance is often shown to be useful in practice~\autocite{lerman2007embedding, brown1970migration, kim2018functional}.
We compare the performance of this cosine-based embedding distance against those derived using dot product similarity and euclidean distance.

We compare the performance of the embedding distance to many baselines. 
These include distances derived from simpler embedding approaches, such as Singular Value Decomposition (SVD) and a Laplacian Eigenmap embedding performed on the underlying co-occurrence matrix. 
We also use network-based distances, calculating vectors using a Personalized Page Rank approach and measuring the distance between them using cosine distance and Jensen-Shannon Divergence (see \SI).
Finally, we compare the embedding distnace against embeddings calculated through direct matrix factorization, following the approach that \textit{word2vec} implicitley approximates~\cite{levy2014neural}.
